Fusing Filters with Integer Linear Programming

### 2014-05-11 2014betterfusionforfilters/section/04-ILP.tex
Add notes to ILP section

\TODO{More justification of why we want to use a heavyweight technology like ILP to do the clustering. Use normalizeInc or one like it to argue that plenty of real programs will admit multiple valid clusterings, so we really want to do this via constraints and an objective function. Compare with prior work on typed fusion that only tries to minimise the number of clusters.}
\begin{code}
 normalizeInc :: Array Int -> Array Int
 normalizeInc xs
  = let incs = map  (+1)    us      (A1) (B2)
        sum  = fold (+) 0   us      (A1) (B1)
        norm = map  (/ sum) incs    (A2) (B2)
    in  norm
\end{code}

@@

\TODO{Put this somewhere: Fusion-preventing dependencies are akin to barrier synchronisations in the Bulk Synchronous model of parallel computation. }

@@

\subsection{Function $\to$ graph}
Converting a function in \emph{combinator normal form} to a graph --- in fact, a DAG --- is quite simple.
Each binding becomes a node in the graph.
When a binding references other arrays or scalars, there is an edge between those two nodes.
Edges may be either \emph{fusible} or \emph{fusion-preventing}.
Fusion-preventing edges mean that the entire input node must completely finish its execution before the output node can start executing.
For example, @fold@s consume their entire input array before producing a single result, so any references to folds must be fusion-preventing.
Conversely, @map@s produce output data for every input element, so may be fused.

@@

\subsection{Graph $\to$ ILP}
With the dependency graph now constructed, it must be converted to an integer linear program.
The integer linear program will contain some variables, an objective function, and constraints that the variables must conform to.
An external solver will find and return a variable assignment that minimises the objective function.
With the variable assignment, the graph's nodes are partitioned into a clustering, and then flow fusion\cite{lippmeier2013flow} is used to extract imperative loops for each cluster.

@@

\begin{tabbing}
M   \= MM \= MMMMMMM \= MM \= \kill
$x$   \> @::@  \> $node \times node$ \> $\to$ \> $\mathbb{B}$
\end{tabbing}
The first and most important group is created for every pair of nodes, and has a boolean value indicating whether the two nodes can be fused.
If $x_{ij} = 0$, then $i$ and $j$ are fused together.
While it may seem counterintuitive for 0 to mean fused and 1 unfused, it means the objective function can simply minimise a weighted $x_{ij}$, and also makes later precedence constraints simpler.
The values of these variables are used to construct the partitioning at the end, such that $\forall i,j.\ x_{ij} = 0 \iff cluster(i) = cluster(j)$.

\begin{tabbing}
M   \= MM \= MMMMMMM \= MM \= \kill
$\pi$ \> @::@  \> $node$             \> $\to$ \> $\mathbb{R}$
\end{tabbing}
The second group of variables is used simply to ensure that the resulting clustering is acyclic.
For some node $i$, we associate a real $\pi_i$ such that every node $j$ after $i$ has a value $\pi_j > \pi_i$.
This has the effect of requiring an acyclic clustering, as if one node is after another its $\pi$ must be strictly greater than its predecessor, and the predecessor must be strictly less than the successor.
If two nodes are to be fused together, their $\pi$ values must be the same.

\begin{tabbing}
M   \= MM \= MMMMMMM \= MM \= \kill
$c$   \> @::@  \> $node$             \> $\to$ \> $\mathbb{B}$
\end{tabbing}
The third, and final, group of variables are purely for minimisation, indicating whether a node's array is \emph{fully contracted}.
They are not required for correctness of the clustering, but are used in the objective function.
A node's array will be fully contracted --- that is, it will be replaced by a scalar --- if all nodes that use this array are fused together:
$c_i = 0 \iff \forall (i,j) \in E. x_{ij} = 0$.

\subsubsection{Unoptimised version}
Before showing the optimised version with certain constraints removed (\S\ref{s:OptimisedConstraints}), this simpler, unoptimised version is shown.
The only difference is that fewer constraints and variables are required in the optimised version, but both versions are equivalent.

@@

Whether there is an edge between $i$ and $j$ or not, if the two are fused together, then $x_{ij} = 0$.
If $x_{ij} = 0$, then both cases may be simplified to $0 \le \pi_j - \pi_i \le 0$, which is equivalent to $\pi_i = \pi_j$.

@@

\paragraph{Fusion between different types}

\ben{We haven't said what a ``type'' is in the context of clustering. Do we mean ``rate''? If so, can we use $k$ for the variables instead of $\tau$?}

\begin{tabbing}
MMMMM   \= MMM \= M \= MMM \= M \= MMM \= \kill
Minimise   \> \ldots \\
Subject to \> \ldots \\
           \> $x_{ij}$    \> $=$   \> $1$             \>       \>            \\
           \> (if $\tau_i \not= \tau_j \wedge trans_i=\emptyset \wedge trans_j=\emptyset$)                 \\
           \> $x_{i'j}$   \> $\le$ \> $x_{ij}$        \>       \>            \\
           \> (if $\tau_i \not= \tau_j \wedge trans_i=\{i'\}$) \\
           \> $x_{ij'}$   \> $\le$ \> $x_{ij}$        \>       \>            \\
           \> (if $\tau_i \not= \tau_j \wedge trans_j=\{j'\}$) \\
           \> \ldots
\end{tabbing}
The main difference to existing integer linear programming solutions is that, while nodes with different loop sizes cannot generally be fused together, we do support some fusion of filtered sizes.
If two nodes have different sizes or rates, but are both sub-rates of another rate, they \emph{can} be fused together if they are both fused with their generators. \ben{What is a subrate? How do we determine the subrates of each rate?}
\TODO{Expand}.

\paragraph{Array contraction}
\begin{tabbing}
MMMMM   \= MMM \= M \= MMM \= M \= MMM \= \kill
Minimise   \> \ldots \\
Subject to \> \ldots \\
           \> $x_{ij}$    \> $\le$ \> $c_i$           \>       \>            \\
           \> (for all edges from i)            \\
           \> \ldots \\
\end{tabbing}
As per Darte's work on optimising for array contraction\cite{darte2002contraction}, we define a variable $c_i$ for each array.
An array is fully contracted if all its output edges are fused with it.
Thus, $c_i=0$ only if $\forall j | (i,j) \in E. x_{ij} = 0$.
By minimising $c_i$ in the objective function, we favour solutions that reduce the number of required intermediate arrays.

>>>>

\TODO{More justification of why we want to use a heavyweight technology like ILP to do the clustering. Use normalizeInc or one like it to argue that plenty of real programs will admit multiple valid clusterings, so we really want to do this via constraints and an objective function. Compare with prior work on typed fusion that only tries to minimise the number of clusters.} \TODO{Also give a quick rundown of what ILP actually is.}
\begin{code}
 normalizeInc :: Array Int -> Array Int
 normalizeInc xs
  = let incs = map  (+1)    us      (A1) (B2)
        sum  = fold (+) 0   us      (A1) (B1)
        norm = map  (/ sum) incs    (A2) (B2)
    in  norm
\end{code}

@@

@@

% -----------------------------------------------------------------------------
\subsection{Function $\to$ graph}
\ben{This is a different sort of graph than before. This is a dependency graph where the nodes are intermediate variables, verses the previous ``program graph'' where the nodes can also be operators. We need to contrast these properly and add a diagram of a dependency graph.}
\TODO{Put this somewhere: Fusion-preventing dependencies are akin to barrier synchronisations in the Bulk Synchronous model of parallel computation. }

Converting a function in \emph{combinator normal form} to a graph --- in fact, a DAG --- is quite simple.
Each binding becomes a node in the graph. When a binding references other arrays or scalars, there is an edge between those two nodes. Edges may be either \emph{fusible} or \emph{fusion-preventing}. Fusion-preventing edges mean that the entire input node must completely finish its execution before the output node can start executing. For example, @fold@s consume their entire input array before producing a single result, so any references to folds must be fusion-preventing. Conversely, @map@s produce output data for every input element, so may be fused.

@@

% -----------------------------------------------------------------------------
\subsection{Graph $\to$ ILP}
With the dependency graph now constructed, it must be converted to an integer linear program. The integer linear program will contain some variables, an objective function, and constraints that the variables must conform to. An external solver will find and return a variable assignment that minimises the objective function. With the variable assignment, the graph's nodes are partitioned into a clustering, and then flow fusion\cite{lippmeier2013flow} is used to extract imperative loops for each cluster.

@@

\begin{tabbing}
M   \= MM \= MMMMMMM \= MM \= \kill
$x$   \> @:@  \> $node \times node$ \> $\to$ \> $\mathbb{B}$
\end{tabbing}
The first and most important group is created for every pair of nodes, and has a boolean value indicating whether the two nodes can be fused. If $x_{ij} = 0$, then $i$ and $j$ are fused together. While it may seem counterintuitive for 0 to mean fused and 1 unfused, it means the objective function can simply minimise a weighted $x_{ij}$, and also makes later precedence constraints simpler. The values of these variables are used to construct the partitioning at the end, such that $\forall i,j.\ x_{ij} = 0 \iff cluster(i) = cluster(j)$. \ben{When we're talking about the dependency graph, when we say ``nodes can be fused'', it means that the operators that produce these values can be fused -- or that the values are produced by the same cluster of operators.}

\begin{tabbing}
M   \= MM \= MMMMMMM \= MM \= \kill
$\pi$ \> @:@  \> $node$             \> $\to$ \> $\mathbb{R}$
\end{tabbing}
The second group of variables is used to ensure that the clustering is acyclic. For each node $i$, we associate a real $\pi_i$ such that every node $j$ after $i$ we have $\pi_j > \pi_i$. This has the effect of requiring an acyclic clustering, as if one node is after another its $\pi$ must be strictly greater than its predecessor, and the predecessor must be strictly less than the successor. If two nodes are to be fused together, their $\pi$ values must be the same. \ben{I don't understand the ``acyclic'' part. If we don't add these constraints, then how might the clustering turn out to be ``cyclic''?}

\begin{tabbing}
M   \= MM \= MMMMMMM \= MM \= \kill
$c$   \> @:@  \> $node$             \> $\to$ \> $\mathbb{B}$
\end{tabbing}
The third, and final, group of variables are purely for minimisation, indicating whether a node's array is \emph{fully contracted}.
They are not required for correctness of the clustering, but are used in the objective function.
A node's array will be fully contracted --- that is, it will be replaced by a scalar --- if all nodes that use this array are fused together: $c_i = 0 \iff \forall (i,j) \in E. x_{ij} = 0$. \ben{Using $\forall$ like this implies that the quantified $i$ isn't the same as the one attached to $c_i$. Write this with a list comprehension instead.}

% -----------------------------------------------------------------------------
\subsubsection{Unoptimised version}
Before showing the optimised version with certain constraints removed (\S\ref{s:OptimisedConstraints}), this simpler, unoptimised version is shown. The only difference is that fewer constraints and variables are required in the optimised version, but both versions are equivalent. \ben{What does ``equivalent'' mean when the constraints are different?} \ben{Say what ``optimised'' means in this case.} \ben{Give some hints about why the unoptimised version has more constraints. }

@@

Whether there is an edge between $i$ and $j$ or not, if the two are fused together, then $x_{ij} = 0$. If $x_{ij} = 0$, then both cases may be simplified to $0 \le \pi_j - \pi_i \le 0$, which is equivalent to $\pi_i = \pi_j$. \ben{Highlight that ``fused together'' means ``in the same cluster'', and that operators can be fused even if there are no direct edges between them.}

@@

\paragraph{Fusion between different types}
\ben{We haven't said what a ``type'' is in the context of clustering. Do we mean ``rate''? If so, can we use $k$ for the variables instead of $\tau$?}

\begin{tabbing}
MMMMM   \= MMM \= M \= MMM \= M \= MMM \= \kill
Minimise   \> \ldots \\
Subject to \> \ldots \\
           \> $x_{ij}$    \> $=$   \> $1$             \>       \>            \\
           \> (if $\tau_i \not= \tau_j \wedge trans_i=\emptyset \wedge trans_j=\emptyset$)                 \\
           \> $x_{i'j}$   \> $\le$ \> $x_{ij}$        \>       \>            \\
           \> (if $\tau_i \not= \tau_j \wedge trans_i=\{i'\}$) \\
           \> $x_{ij'}$   \> $\le$ \> $x_{ij}$        \>       \>            \\
           \> (if $\tau_i \not= \tau_j \wedge trans_j=\{j'\}$) \\
           \> \ldots
\end{tabbing}
The main difference to existing integer linear programming solutions is that, while nodes with different loop sizes cannot generally be fused together, we do support some fusion of filtered sizes. If two nodes have different sizes or rates, but are both sub-rates of another rate, they \emph{can} be fused together if they are both fused with their generators. \ben{What is a subrate? How do we determine the subrates of each rate?} \ben{Highlight why we want to allow this sort of fusion, and why it is specific to data flow fusion.} \ben{Explain how we are using the idea of transducers}.

\paragraph{Array contraction}
\begin{tabbing}
MMMMM   \= MMM \= M \= MMM \= M \= MMM \= \kill
Minimise   \> \ldots \\
Subject to \> \ldots \\
           \> $x_{ij}$    \> $\le$ \> $c_i$           \>       \>            \\
           \> (for all edges from i)            \\
           \> \ldots \\
\end{tabbing}
As per Darte's work on optimising for array contraction\cite{darte2002contraction}, we define a variable $c_i$ for each array.
An array is fully contracted if all its output edges are fused with it. Thus, $c_i=0$ only if $\forall j | (i,j) \in E. x_{ij} = 0$. By minimising $c_i$ in the objective function, we favour solutions that reduce the number of required intermediate arrays. \ben{Say what array contraction is, and why we want this constraint.}
